Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-07-24 analysis/paper/paper.Rmd
added sentences about keywords in GMM in conclusion

Indigenous societies' responses to colonial contact ranges from passive acceptance to active negotiation with the colonists, and accommodation or resistance of foreign intrusion, which can be identified through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (ca. AD 1350-1850), a large multi-component archaeological site in Yilan County, northeastern Taiwan [@Chen2007], to identify the indirect impacts of colonial settler activity on local Indigenous societies. Yilan is an ideal context to study peripheral colonial influences because the Indigenous communities were isolated by geographical barriers, limiting the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing east toward the Pacific with high mountains on three other sides.

This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in local Indigenous societies. We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization caused by greater economic and social control of ceramic production by a small group of individuals. Using standardization in ceramic shapes as a proxy for craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels? Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?

@@

Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark and outline approaches [@Adams2004]. Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017]. With univariate or multivariate statistics for shape data generated by GMM, the application to artifacts can answer research questions related to topological evolution or variations in assemblages to define groups [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova]. A common procedure is using dimensional reduction techniques, such as principal components analysis or canonical variate analysis, to capture the essential features that can represent the overall shape. This enables the identification of clustered analytical groups through the visualization of results, followed by statistical tests to robustly distinguish them.

@@

<!-- size analysis -->

We also conducted size analysis by examining the body diameter of the vessels to explore any differences in vessel form reflected by size and shape. Body diameter is a suitable metric measurement to represent size for ceramics since it is associated with volume as a utilitarian function, and could present differences in the situation of similar shape. The body diameter of the vessels from Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-regression): A). It is interesting that linear regression model (Figure \@ref(fig:size-regression): B) predicts that shape, captured by PC1 and PC2, changes with size across three phases, except PC1 in Chinese period. That is, the shorter vessels are larger in body diameter according to the significant positive correlation between PC1 and size; however, the vessel from Chinese period does not show this pattern. On the contrary, the negative relationship between PC2 and size suggests that the vessels with narrower neck and mouth tend to have a larger body diameter.

```{r size-regression, fig.cap= "A: Distribution of the body diameter of the vessels across three phases. B: Scatter plots of shape variable, PC1 at the top and PC2 at the bottom, as a function of body diamater"}

@@

# can pc1 predict size by period?
pc1_size_reg <- 
  ggscatter(jpgs_imported_coo_size_pcs_metric, 
            x = "Diameter1_Body", 
            y = "PC1",
            add = "reg.line",
            size = 1) +
    facet_wrap(~ fct_rev(y)) +
    stat_cor(label.y = 1.25, label.sep = "\n") +
    #stat_regline_equation(label.y = 1) +
    labs(x = "") +
    theme(axis.text = element_text(size = 8)) +
    #scale_x_continuous(breaks = c(75, 200)) +
    scale_y_continuous(limits = c(0, 1.5),
                       breaks = c(seq(0, 1, 0.5)),
                       labels = c(seq(0, 1, 0.5)))

# can pc2 predict size by period?
pc2_size_reg <- 
  ggscatter(jpgs_imported_coo_size_pcs_metric, 
            x = "Diameter1_Body", 
            y = "PC2",
            add = "reg.line",
            size = 1) +
    facet_wrap(~ fct_rev(y)) +
    stat_cor(label.y = 1.25, label.sep = "\n") +
    #stat_regline_equation(label.y = 1) +
    labs(x = "body diameter (mm)") +
    theme(axis.text = element_text(size = 8)) +
    #scale_x_continuous(breaks = c(75, 200)) +
    scale_y_continuous(limits = c(0, 1.5),
                       breaks = c(seq(0, 1, 0.5)),
                       labels = c(seq(0, 1, 0.5)))

@@

@@

This study demonstrated an application of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences. Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust method of identifying and statistically assessing differences in shape standardization. The direct relationship between foreign influences and standardization of ceramic shape was tested on ceramics from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan. Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence. Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context. More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggest that shape standardization was intentional. The results further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts. Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.

>>>>

Indigenous societies' responses to colonial contact ranges from passive acceptance to active negotiation with the colonists, and accommodation or resistance of foreign intrusion, which can be identified through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. In this paper we investigate the archaeology of a pericolonial situation at Kiwulan (ca. AD 1350-1850), a large multi-component archaeological site in Yilan County, northeastern Taiwan [@Chen2007], to identify the indirect impacts of colonial settler activity on local Indigenous societies. Yilan is an ideal context to study peripheral colonial influences because the Indigenous communities were isolated by geographical barriers, limiting the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan County, which is characterized by a triangular alluvial plain facing east toward the Pacific with high mountains on three other sides. 
This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in local Indigenous societies. We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization caused by greater economic and social control of ceramic production by a small group of individuals. Using standardization in ceramic shapes as a proxy for craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels? Did pottery shape become more homogeneous and standardized after foreign contacts with European colonizers or Chinese immigrants?

@@

Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark and outline approaches [@Adams2004]. Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017]. With univariate or multivariate statistics, the GMM applications can answer research questions related to lithic topological evolution and variations [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019]. A common procedure is using dimensional reduction techniques, such as principal components analysis or canonical variate analysis, to capture the essential features that can represent the overall shape. This enables the identification of clustered analytical groups through the visualization of results, followed by statistical tests to robustly distinguish them.

@@

<!-- size analysis -->
We also conducted size analysis by examining the body diameter of the vessels to explore any differences in vessel form reflected by size and shape. Body diameter is a suitable metric measurement to represent size for ceramics since it is associated with volume as a utilitarian function, and could present differences in the situation of similar shape. The body diameter of the vessels from Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A). It is interesting that linear regression model (Figure \@ref(fig:size-shape-regression-plots): B) predicts that shape, captured by PC1 and PC2, changes with size across three phases, except PC1 in Chinese period. That is, the shorter vessels are larger in body diameter according to the significant positive correlation between PC1 and size; however, the vessel from Chinese period does not show this pattern. On the contrary, the negative relationship between PC2 and size suggests that the vessels with narrower neck and mouth tend to have a larger body diameter.

```{r size-analysis}

@@

# can pc1 predict size by period?
pc1_size_reg <- 
  ggscatter(jpgs_imported_coo_size_pcs_metric, 
            x = "Diameter1_Body", 
            y = "pc1",
            add = "reg.line",
            size = 1) +
    facet_wrap(~ fct_rev(y)) +
    stat_cor(label.y = 1.25, label.sep = "\n") +
    #stat_regline_equation(label.y = 1) +
    labs(x = "", y = "PC1") +
    theme(axis.text = element_text(size = 8)) +
    #scale_x_continuous(breaks = c(75, 200)) +
    scale_y_continuous(limits = c(0, 1.5),
                       breaks = c(seq(0, 1, 0.5)),
                       labels = c(seq(0, 1, 0.5)))

# can pc2 predict size by period?
pc2_size_reg <- 
  ggscatter(jpgs_imported_coo_size_pcs_metric, 
            x = "Diameter1_Body", 
            y = "pc2",
            add = "reg.line",
            size = 1) +
    facet_wrap(~ fct_rev(y)) +
    stat_cor(label.y = 1.25, label.sep = "\n") +
    #stat_regline_equation(label.y = 1) +
    labs(x = "body diameter (mm)", y = "PC2") +
    theme(axis.text = element_text(size = 8)) +
    #scale_x_continuous(breaks = c(75, 200)) +
    scale_y_continuous(limits = c(0, 1.5),
                       breaks = c(seq(0, 1, 0.5)),
                       labels = c(seq(0, 1, 0.5)))

@@

```{r, size-shape-regression-plots, echo=FALSE, fig.cap= "A: Distribution of the body diameter of the vessels across three phases. B: Scatter plots of shape variable, PC1 at the top and PC2 at the bottom, as a function of body diamater"}
knitr::include_graphics(here("analysis/figures/size-boxplot-regression.jpg"))
```

@@

This study demonstrated an application of EFA on ceramic shapes to explore the emergence of ceramic specialization as indicative of foreign influences. Here, EFA is combined with a significance test for the equality of CVs of shape variables to provide a robust method of identifying and statistically assessing differences in shape standardization. The direct relationship between foreign influences and standardization of ceramic shape was tested on ceramics from Kiwulan, a large Iron Age Indigenous settlement in northeastern Taiwan. Lower variation in ceramic shape was identified after European presence began, and even lower variation during the period of Chinese presence. Our findings help to expand upon those factors that may lead to the standardization of ceramic production in a pericolonial interaction context. More homogeneous shapes during the contact periods, without any changes in clay paste composition or production technique, suggest that shape standardization was intentional. The results further suggest that expressions of social identity or cultural boundaries in Indigenous societies through highly visible vessel qualities, such as shape, may be heightened during periods of foreign contact in pericolonial contexts. Current GMM method are mostly applied to study lithics topology, variations in bone shape in the filed of bioarchaeology or zooarchaeology, and some new attempts for understanding cutmarks and tophomic processes. We demonstrate that how we can use EFA-based GMM to identify the differences in ceramics to tackle the issue of specialized production. Our analysis, with its openly available methods and data, is readily extensible to other pottery assemblages in the region to further explore related questions about craft specialization and standardization in Iron Age ceramic technologies.
